The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2012-07-03 script/de/chapters/introduction.tex
[i18n] Small update in CT script (starting with CT2 template tree)

\vspace{-12pt}
Ab dem Jahr 2000 wurde mit dem CrypTool-Paket\index{CrypTool} auch ein
Skript ausgeliefert, das die Mathematik einzelner Themen genauer, aber doch 
möglichst verständlich erläutern sollte.

Um auch hier die Möglichkeit zu schaffen, dass getrennte Entwickler/Autoren
mitarbeiten können, wurden die Themen sinnvoll unterteilt und dafür jeweils
eigenständig lesbare Kapitel geschrieben. In der anschließenden
redaktionellen Arbeit wurden in TeX Querverweise ergänzt und Fußnoten
hinzugefügt, die zeigen, an welchen Stellen man die entsprechenden
Funktionen in CrypTool v1\index{CrypTool 1.x} aufrufen kann 
\hyperlink{appendix-menutree}{(vgl. den Menübaum} im Anhang \ref{s:appendix-menutree}).
%\hypertarget{appendix-menutree}{}\label{s:appendix-menutree}
Natürlich gäbe es viel mehr Themen in Mathematik und Kryptographie, die
man vertiefen könnte -- deshalb ist diese Auswahl auch nur eine von
vielen möglichen.

@@

Dieses Skript wird zusammen mit dem Open-Source-Programm CrypTool\index{CrypTool}
ausgeliefert.

@@

Während für das Verständnis der meisten Kapitel Abiturwissen ausreicht,
erfordern Kapitel \ref{Chapter_ModernCryptography} (Moderne Kryptografie)
und \ref{Chapter_EllipticCurves} (Elliptische Kurven) tiefere mathematische
Kenntnisse.

@@

\noindent \textbf{Das Programm CrypTool\index{CrypTool}}

CrypTool\index{CrypTool} ist ein Lernprogramm mit umfangreicher Online-Hilfe,
mit dem Sie unter einer einheitlichen Oberfläche
kryptographische Verfahren anwenden und analysieren können.

@@

An dieser Stelle möchte ich explizit folgenden Personen danken, 
die bisher in ganz besonderer Weise zu CrypTool\index{CrypTool} 
beigetragen haben.
Ohne ihre besonderen Fähigkeiten und ihr großes Engagement wäre CrypTool
nicht, was es heute ist:
\vspace{-7pt}
\begin{list}{\textbullet}{\addtolength{\itemsep}{-0.5\baselineskip}}
   \item Hr.\ Henrik Koy
   \item Hr.\ Jörg-Cornelius Schneider
   \item Hr.\ Florian Marchal
   \item Dr.\ Peer Wichmann
   \item Mitarbeiter von Prof.\ Claudia Eckert, Prof.\ Johannes Buchmann und Prof.\ Torben Weis.
\end{list}
Auch allen hier nicht namentlich Genannten ganz herzlichen Dank für das 
(meist in der Freizeit) geleistete Engagement.
\\
\\
Bernhard Esslinger
\\
\\
Frankfurt, November 2009

>>>>

\vspace{-12pt}
Ab dem Jahr 2000 wurde mit dem CrypTool-1-Paket\index{CrypTool} auch ein
Skript ausgeliefert, das die Mathematik einzelner Themen genauer, aber doch 
möglichst verständlich erläutern sollte.

Um auch hier die Möglichkeit zu schaffen, dass getrennte Entwickler/Autoren
mitarbeiten können, wurden die Themen sinnvoll unterteilt und dafür jeweils
eigenständig lesbare Kapitel geschrieben. In der anschließenden
redaktionellen Arbeit wurden in TeX Querverweise ergänzt und Fußnoten
hinzugefügt, die zeigen, an welchen Stellen man die entsprechenden
Funktionen in CrypTool 1 (CT1)\index{CrypTool 1.x} aufrufen kann 
\hyperlink{appendix-menutree}{(vgl. den Menübaum} im Anhang \ref{s:appendix-menutree}).
%\hypertarget{appendix-menutree}{}\label{s:appendix-menutree}
Natürlich gäbe es viel mehr Themen in Mathematik und Kryptographie, die
man vertiefen könnte -- deshalb ist diese Auswahl auch nur eine von
vielen möglichen.

@@

Dieses Skript wird zusammen mit dem Open-Source-Programm CrypTool 1\index{CrypTool}
ausgeliefert.

@@

Während für das Verständnis der meisten Kapitel Abiturwissen ausreicht,
erfordern die Kapitel \ref{Chapter_ModernCryptography} (Moderne Kryptografie)
und \ref{Chapter_EllipticCurves} (Elliptische Kurven) tiefere mathematische
Kenntnisse.

@@

\noindent \textbf{Das Programm CrypTool 1\index{CrypTool}}

CrypTool 1 (CT1)\index{CrypTool} ist ein Lernprogramm mit umfangreicher
Online-Hilfe, mit dem Sie unter einer einheitlichen Oberfläche
kryptographische Verfahren anwenden und analysieren können.

@@

An dieser Stelle möchte ich explizit folgenden Personen danken, 
die bisher in ganz besonderer Weise zu CT1\index{CrypTool} 
beigetragen haben.
Ohne ihre besonderen Fähigkeiten und ihr großes Engagement wäre CrypTool
nicht, was es heute ist:
\vspace{-7pt}
\begin{list}{\textbullet}{\addtolength{\itemsep}{-0.5\baselineskip}}
   \item Hr.\ Henrik Koy
   \item Hr.\ Jörg-Cornelius Schneider
   \item Hr.\ Florian Marchal
   \item Dr.\ Peer Wichmann
   \item Mitarbeiter von Prof.\ Claudia Eckert, Prof.\ Johannes Buchmann,
         Prof.\ Torben Weis and Prof.\ Arno Wacker.
\end{list}
Auch allen hier nicht namentlich Genannten ganz herzlichen Dank für das 
(meist in der Freizeit) geleistete Engagement.
\\
\\
Bernhard Esslinger
\\
\\
Frankfurt, Juli 2012

### 2012-07-03 script/de/chapters/menus.tex
[i18n] Small update in CT script (starting with CT2 template tree)

   % Eyecatcher_Neue-CrypTool-Version
Dieser Anhang enthält auf der folgenden Seite den kompletten Menübaum der
CrypTool\index{CrypTool}-Version 1.4.30\footnote{%
  Parallel zu CrypTool 1.x\index{CrypTool 1.x} werden im CrypTool-Projekt
  momentan auch die Zukunftsversionen CrypTool 2\index{CrypTool 2.0} und
  JCrypTool\index{JCrypTool 1.0} entwickelt:\\
  - Webseite CT2: \url{http://www.cryptool2.vs.uni-due.de} \\
  - Webseite JCT: \url{http://jcryptool.sourceforge.net} \\
  Diese Zukunftsversionen sind zur Zeit (Dezember 2009) noch Betaversionen;
  sie sind aber schon stabil genug, um von Endbenutzern genutzt werden
  zu können. Wenn es hier Releaseversionen gibt, werden im Skript die
  entsprechenden Menüpfade und Menübäume ergänzt.
}.

\noindent Das Haupt-Menü enthält die Service-Funktionen in den Menüs
\begin{itemize}
   \item Datei
   \item Bearbeiten
   \item Ansicht
   \item Optionen
   \item Fenster
   \item Hilfe
\end{itemize}
und die eigentlichen Krypto-Funktionen in den Menüs
\begin{itemize}
   \item Ver-/Entschlüsseln
   \item Digitale Signaturen/PKI
   \item Einzelverfahren
   \item Analyse.
\end{itemize}

@@

\clearpage
\begin{figure}[hb]
\begin{center}
\vspace{-30pt}
%\frame{
%\includegraphics[scale=0.25, angle=270, viewport=200 30 2680 1420]
\includegraphics[scale=0.35, angle=270]
                {figures/cryptool-menu-de}
%viewport=rand-links? rand-unten breite hoehe [bezogen auf querformat]
%}
\caption{Komplette Übersicht über den Menü-Baum von CrypTool 1.4.30} 
\label{menuoverview}
\end{center}
\end{figure}
\clearpage

>>>>

   % Eyecatcher_Neue-CrypTool-Version
Dieser Anhang enthält auf der folgenden Seite den kompletten Menübaum der
CrypTool\index{CrypTool}-Version 1.4.31\footnote{%
  Parallel zu CrypTool 1.x (CT1)\index{CrypTool 1.x} werden im CrypTool-Projekt
  momentan auch die Zukunftsversionen CrypTool 2 (CT2)\index{CrypTool 2.0} und
  JCrypTool (JCT)\index{JCrypTool 1.0} entwickelt:\\
  - Webseite CT2: \url{http://www.cryptool2.vs.uni-due.de} \\
  - Webseite JCT: \url{http://jcryptool.sourceforge.net} \\
  Diese Zukunftsversionen sind zur Zeit (Dezember 2009) noch Betaversionen;
  sie sind aber schon stabil genug, um von Endbenutzern genutzt werden
  zu können.
}.

\noindent Das Haupt-Menü von CT1 enthält die Service-Funktionen in den Menüs
\begin{itemize}
   \item Datei
   \item Bearbeiten
   \item Ansicht
   \item Optionen
   \item Fenster
   \item Hilfe
\end{itemize}
und die eigentlichen Krypto-Funktionen in den Menüs
\begin{itemize}
   \item Ver-/Entschlüsseln
   \item Digitale Signaturen/PKI
   \item Einzelverfahren
   \item Analyse.
\end{itemize}

@@

\clearpage
\begin{figure}[hb]
\begin{center}
\vspace{-30pt}
%\frame{
%\includegraphics[scale=0.25, angle=270, viewport=200 30 2680 1420]
\includegraphics[scale=0.35, angle=270]
                {figures/CT1-menutree-de}
%viewport=rand-links? rand-unten breite hoehe [bezogen auf querformat]
%}
\caption{Komplette Übersicht über den Menü-Baum von CT1 (CrypTool 1.4.31)} 
\label{menuoverview}
\end{center}
\end{figure}
\clearpage

\noindent Text zu CT2 xxxxxxxxxx DRAFT

\clearpage
\begin{figure}[hb]
\begin{center}
\vspace{-30pt}
\includegraphics[scale=0.8, angle=270]
                {figures/CT2-templatetree-de}
\caption{Übersicht über den Template-Baum von CT2 (NB4882.1, Juli 2012), Teil 1} 
\label{menuoverview}
\end{center}
\end{figure}
\clearpage

### 2012-07-03 script/en/chapters/introduction.tex
Script[de+en]: Factorisation of RSA-768 added plus some minor things.
[i18n] Small update in CT script (starting with CT2 template tree)

Starting in the year 2000 this script became part of the 
CrypTool v1\index{CrypTool 1.x} package. It is designed to accompany the program 
CrypTool by explaining some mathematical topics in more detail, 
but still in a way which is easy to understand.

In order to also enable developers/authors to work together independently 
the topics have been split up and for each topic an extra chapter has been 
written which can be read on its own. The later editorial work in TeX added 
cross linkages between different sections and footnotes describing where you
can find the according functions within the CrypTool v1\index{CrypTool 1.x} 
program \hyperlink{appendix-menutree}{(see menu tree} in appendix \ref{s:appendix-menutree}).
% \hypertarget{appendix-menutree}{}\label{s:appendix-menutree}
Naturally there are many more interesting topics in mathematics and
cryptography which could be discussed in greater depth -- therefore this
is only one of many ways to do it.

@@

%This edition of the script adds some topics, but mainly updates areas (e.g. the
%summaries of topical research areas):
This edition completely updated the TeX sources of the document, and of course
the content of the script was corrected, amended and updated with some topics, e.g.:
\vspace{-5pt}
\begin{itemize}
  \item the search for the largest prime numbers 
        (chap. \ref{search_for_very_big_primes}), 
  \item progress in cryptanalysis of hash algorithms 
        (chap. \ref{NeueAES-Analyse}) and
%  \item progress in cryptanalysis of hash algorithms 
%        (chap. \ref{collision-attacks-against-sha-1}) and
%  \item progress in ideas for new crypto methods (RSA successor) 
%        (chap. \ref{xxxxxxxxxBrute-force-gegen-Symmetr})\index{xxxxxxxxxxxxxxxx} and
  \item the list of movies or novels, in which cryptography or number theory 
        played major role (see appendix \ref{s:appendix-movies});
        and where primes are used as hangers  
        (see curiouses in \ref{HT-Quaint-curious-Primes-usage}).
  \item Newly added are e.g. the section about Benford's law and primes
        (chap. \ref{primes:Benford-Law}), and 
        the appendix \ref{s:appendix-using-sage} about using the
        computer algebra system Sage. Sage becomes
        more and more the standard open-source CAS system. Accordingly all
        samples written before in Pari-GP and Mathematica have been substituted
        with Sage code. Thanks to Nguyen and Massierer, a lot of new code samples
        could be added (see also chap.~\ref{ec:Sage_Massierer}).
\end{itemize}

@@

This document is delivered together with the open-source program CrypTool\index{CrypTool}.

@@

\noindent \textbf{The program CrypTool\index{CrypTool}}

CrypTool\index{CrypTool} is an educational program with a comprehensive online
help enabling you to use and analyse cryptographic procedures within a
unified graphical user interface.

@@

At this point I'd like to thank explicitly the following people who
particularly contributed to CrypTool\index{CrypTool}. They applied their
very special talents and showed really great engagement:
\vspace{-7pt}
%\begin{itemize}
\begin{list}{\textbullet}{\addtolength{\itemsep}{-0.5\baselineskip}}
   \item Mr.\ Henrik Koy
   \item Mr.\ J\"org-Cornelius Schneider
   \item Mr.\ Florian Marchal
   \item Dr.\ Peer Wichmann
   \item Staff of Prof.\ Claudia Eckert, Prof.\ Johannes Buchmann and Prof.\ Torben Weis.
\end{list}
%\end{itemize}
Also I want to thank all the many people not mentioned here for their 
hard work (mostly carried out in their spare time).
\\
\\
Bernhard Esslinger
\\
\\
Frankfurt (Germany), November 2009

>>>>

Starting in the year 2000 this script became part of the 
CrypTool 1 (CT1)\index{CrypTool 1.x} package. It is designed to accompany the
program CrypTool by explaining some mathematical topics in more detail, 
but still in a way which is easy to understand.

In order to also enable developers/authors to work together independently 
the topics have been split up and for each topic an extra chapter has been 
written which can be read on its own. The later editorial work in TeX added 
cross linkages between different sections and footnotes describing where you
can find the according functions within the CrypTool 1\index{CrypTool 1.x} 
program \hyperlink{appendix-menutree}{(see menu tree} in appendix \ref{s:appendix-menutree}).
% \hypertarget{appendix-menutree}{}\label{s:appendix-menutree}
Naturally there are many more interesting topics in mathematics and
cryptography which could be discussed in greater depth -- therefore this
is only one of many ways to do it.

@@

%This edition of the script adds some topics, but mainly updates areas (e.g. the
%summaries of topical research areas):
This edition completely updated the TeX sources of the document, and of course
the content of the script was corrected, amended and updated with some topics, e.g.:
\vspace{-5pt}
\begin{itemize}
  \item the largest prime numbers (chap. \ref{search_for_very_big_primes}),
        new factorization records (chap. \ref{RSA-768}),
  \item progress in cryptanalysis of AES 
        (chap. \ref{NeueAES-Analyse}) and
%  \item progress in cryptanalysis of hash algorithms 
%        (chap. \ref{collision-attacks-against-sha-1}) and
%  \item progress in ideas for new crypto methods (RSA successor) 
%        (chap. \ref{xxxxxxxxxBrute-force-gegen-Symmetr})\index{xxxxxxxxxxxxxxxx} and
  \item the list of movies or novels, in which cryptography or number theory 
        played major role (see appendix \ref{s:appendix-movies});
        and where primes are used as hangers  
        (see curiouses in \ref{HT-Quaint-curious-Primes-usage}).
  \item Newly added are e.g. the section about Benford's law and primes
        (chap. \ref{primes:Benford-Law}), and 
        the appendix \ref{s:appendix-using-sage} about using the
        computer algebra system Sage. Sage becomes
        more and more the standard open-source CAS system. Accordingly all
        samples written before in Pari-GP and Mathematica have been substituted
        with Sage code. Thanks to Nguyen and Massierer, a lot of new code
        samples could be added (see also chap.~\ref{ec:Sage_Massierer}).
\end{itemize}

@@

This document is delivered together with the open-source program CrypTool 1\index{CrypTool}.

@@

\noindent \textbf{The program CrypTool 1\index{CrypTool}}

CrypTool 1 (CT1)\index{CrypTool} is an educational program with a comprehensive
online help enabling you to use and analyse cryptographic procedures within a
unified graphical user interface.

@@

At this point I'd like to thank explicitly the following people who
particularly contributed to CT1\index{CrypTool}. They applied their
very special talents and showed really great engagement:
\vspace{-7pt}
%\begin{itemize}
\begin{list}{\textbullet}{\addtolength{\itemsep}{-0.5\baselineskip}}
   \item Mr.\ Henrik Koy
   \item Mr.\ J\"org-Cornelius Schneider
   \item Mr.\ Florian Marchal
   \item Dr.\ Peer Wichmann
   \item Staff of Prof.\ Claudia Eckert, Prof.\ Johannes Buchmann,
         Prof.\ Torben Weis and Prof.\ Arno Wacker..
\end{list}
%\end{itemize}
Also I want to thank all the many people not mentioned here for their 
hard work (mostly carried out in their spare time).
\\
\\
Bernhard Esslinger
\\
\\
Frankfurt (Germany), July 2012

### 2012-07-03 script/en/chapters/menus.tex
updated CrypTool menu tree
[i18n] Small update in CT script (starting with CT2 template tree)

This appendix contains at the following page the complete menu tree of
CrypTool\index{CrypTool} version 1.4.30\footnote{%
  In parallel to CrypTool 1.x\index{CrypTool 1.x} the future versions
  CrypTool 2\index{CrypTool 2.0} and JCrypTool\index{JCrypTool 1.0}
  are currently developed in the CrypTool project.\\
  - Webseite CT2: \url{http://www.cryptool2.vs.uni-due.de} \\
  - Webseite JCT: \url{http://jcryptool.sourceforge.net} \\
  These future versions are currently (December 2009) beta versions, but they
  are stable enough to be used by end-users already. If the according release
  versions are available, we will add the appropriate menus and menu trees.\\
}.

\noindent The main menu contains service functions in the menus
\begin{itemize}
   \item File
   \item Edit
   \item View
   \item Options
   \item Window
   \item Help
\end{itemize}
and the actual cyrpto functions in the menus
\begin{itemize}
   \item Crypt/Decrypt
   \item Digital Signature/PKI
   \item Individual Procedures
   \item Analysis.
\end{itemize}

Below \verb#Individual Procedures# you find visualizations of single algorithms
and of protocols. Some procedures are implemented both for a fast performance
(mostly under the main menu \verb#Crypt/Decrypt#) and for a step-by-step visualization.

@@

%\nobreak
\clearpage
\begin{figure}[hb]
\begin{center}
\vspace{-30pt}
%\frame{  %TeX macht einen Rahmen (siehe viewport) drumherum -- gut zum Testen
%\includegraphics[scale=0.25, angle=270, viewport=200 30 2660 1430]
\includegraphics[scale=0.35, angle=270]
                {figures/cryptool-menu-en}
%viewport=rand-links? rand-unten breite hoehe? [bezogen auf querformat]
%}
\caption{Complete overview of the menu tree of CrypTool 1.4.30} 
\label{menuoverview}
\end{center}
\end{figure}
\clearpage

>>>>

This appendix contains at the following page the complete menu tree of
CrypTool\index{CrypTool} version 1.4.31\footnote{%
  In parallel to CrypTool 1.x (CT1)\index{CrypTool 1.x} the future versions
  CrypTool 2 (CT2)\index{CrypTool 2.0} and JCrypTool (JCT)\index{JCrypTool 1.0}
  are currently developed in the CrypTool project.\\
  - Webseite CT2: \url{http://www.cryptool2.vs.uni-due.de} \\
  - Webseite JCT: \url{http://jcryptool.sourceforge.net} \\
  These future versions are currently (July 2012) beta versions, but they
  are stable enough to be used by end-users already.\\
}.

\noindent The main menu of CT1 contains service functions in the menus
\begin{itemize}
   \item File
   \item Edit
   \item View
   \item Options
   \item Window
   \item Help
\end{itemize}
and the actual cyrpto functions in the menus
\begin{itemize}
   \item Encrypt/Decrypt
   \item Digital Signature/PKI
   \item Individual Procedures
   \item Analysis.
\end{itemize}

Below \verb#Individual Procedures# you find visualizations of single algorithms
and of protocols. Some procedures are implemented both for a fast performance
(mostly under the main menu \verb#Encrypt/Decrypt#) and for a step-by-step visualization.

@@

%\nobreak
\clearpage
\begin{figure}[hb]
\begin{center}
\vspace{-30pt}
%\frame{  %TeX macht einen Rahmen (siehe viewport) drumherum -- gut zum Testen
%\includegraphics[scale=0.25, angle=270, viewport=200 30 2660 1430]
\includegraphics[scale=0.35, angle=270]
                {figures/CT1-menutree-en}
%viewport=rand-links? rand-unten breite hoehe? [bezogen auf querformat]
%}
\caption{Complete overview of the menu tree of CT1 (CrypTool 1.4.31)} 
\label{menuoverview}
\end{center}
\end{figure}
\clearpage

\noindent Text for CT2 xxxxxxxxxx DRAFT

\clearpage
\begin{figure}[hb]
\begin{center}
\vspace{-30pt}
\includegraphics[scale=0.8, angle=270]
                {figures/CT2-templatetree-en}
\caption{Overview of the template tree of CT2 (NB4882.1, July 2012), Part 1} 
\label{menuoverview}
\end{center}
\end{figure}
\clearpage
